---
title: "Introduction to Data Visualization with R"
author: "Martin Barron"
date: "October 24, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Setup

1. Please visit [link](https://github.com/m-barron/data_visualization_workshop) to download the workshop materials.
2. Extract the zip file.
3. Start up R-studio.
4. Install necessary packages:

```{r Install Packages, eval=FALSE}
install.packages("tidyverse") # Installs ggplot2 + some other useful packages
library(tidyverse) # Loads ggplot2 + some other useful packages
```

## Workshop Structure:

* Each core concept in ggplot2 will be introduced with examples. 
* There will be short exercises after each concept. Collaboration is encouraged when working on the exercises. 

This is an introductory data visualization workshop:

* R itself will not be taught, though feel free to ask for help.
* We will focus on the `ggplot2` package. Other packages will not be covered.  


## Goal 

```{r goal plot, echo=FALSE, warning=FALSE}
options(scipen=999)  # turn-off scientific notation like 1e+48
library(ggplot2)
data("midwest", package = "ggplot2")
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + 
  xlim(c(0, 0.1)) + 
  ylim(c(0, 500000)) + 
  labs(subtitle="Area Vs Population", 
       y="Population", 
       x="Area", 
       title="Scatterplot", 
       caption = "Source: midwest") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme_bw()

plot(gg)
```


## `ggplot2` 

`ggplot2` entails a grammer of graphics. That is, every graph is built from the same few parts.

## Strengths of `ggplot2`

1. It is easy to get a good looking plot.
2. Very flexible. 
3. A lot of the data processing is taken care of behind the scenes.
4. There is a clear syntax. 
5. Multi-dimensional approach is straight forward allowing many variables to be visualized at once. 
6. The default colors are equally spaced on the color wheel. 

## `ggplot2` does not do
 
1. 3-dimensional graphics - `rgl`
2. Graph theory graphs - `igraph`
3. Interactive graphics - `plotly`

## Key ideas of ggplot2:

1. Aesthetics 
2. Layers & geoms (geometric objects)
3. Facets
4. Themes

## Aesthetics 

An aesthetic is something you map data to and can see in `ggplot`, including:

* Position (X and Y axis).
* Color (Outside color).
* Fill (Inside color).
* Shape of points.
* Line type.
* Size of points. 

## Simple Plot

```{r plot 2}
ggplot(midwest, aes(x=area, y=poptotal)) # Create ggplot2, 

```


## Layers and Geoms 

Geoms are geometric objects, the actual marks on the plot, which include:

* Points - (`geom_point`) For scatter plots, dot plots, ...
* Lines - (`geom_line`) For time series, trend lines, ...
* Boxplots - (`geom_boxplot`)
* and many more.

## Scatter Plot 


```{r Scatter Plot}
ggplot(midwest, aes(x=area, y=poptotal)) + # Create ggplot2, x-axis area, y-axis poptotal
  geom_point() # Use geom point to add points
```

## Pipes

To add elements to a `ggplot` we use `+` at the end of each line. This is called a pipe, which is simply an operator that sends what comes before to what comes after. 

## Regression Line (Multiple Geoms)

```{r regression line}
midwest$pop_pred <- predict(lm(poptotal ~ area, data=midwest))
ggplot(midwest, aes(x=area, y=poptotal)) + # Create ggplot2, x-axis area, y-axis poptotal
  geom_point() + # Use geom point to add points
  geom_line(aes(y = pop_pred))
```


## Smoothers


```{r Add smooth}
ggplot(midwest, aes(x=area, y=poptotal)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() + # Use geom point to add points.
  geom_smooth(method="lm") # Use geom_smooth to add a line of best fit.
```

## Assign Plots

```{r assign plot}

g_1 <- ggplot(midwest, aes(x=area, y=poptotal)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() + # Use geom point to add points.
  geom_smooth(method="lm") # Use geom_smooth to add a line of best fit.
```

## Assign Plots

```{r plot objects}
g_1
```

## Add text (labelling points)

```{r IL plot}
g_2 <- ggplot(subset(midwest, state %in% "OH"), # Select subset of data, only OH
              aes(x=area, y=percbelowpoverty)) + # Create ggplot2, x-axis popadults, y-axis poptotal.
  geom_point() # Use geom point to add points.
g_2
```

## Add text

```{r Label Plot}
g_3 <- g_2 + geom_text(aes(label=county),size=2)
g_3
```

## `ggrepel`

```{r ggrepel}
#install.packages("ggrepel")
library("ggrepel")
g_2 + 
  geom_text_repel(aes(label=county), size = 2)
```

## Mapping variables to other aesthetics

```{r other mappings}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() # Use geom point to add points.
```

## Aesthetic mapping v aesthetic assignment


```{r set aesthetics}
ggplot(midwest, aes(x=area, y=poptotal ,shape=state)) + 
  geom_point(color="blue") # Use geom point to add points.
```

## Exercises 1 

* Create a scatter plot with percent college educated, "percollege", on the X axis and percent below the poverty line, "percbelowpoverty" on Y axis. 
* Color the points red
* Map the color of the points to population density, "popdensity"
* Map the shape of the points to the state, "state"
* Make the points bigger by setting size to "3"
* Map the size of the points to total population, "poptotal"
* Add a trend line to the plot


## Statistical Transformation

While some plot types (e.g. scatter plots) do not require statistical transformations - i.e. each point is plotted at the X and Y coordinates corresponding to the original value. Many other plot types (e.g. box plots, histograms, prediction lines etc.) require the data to be transformed prior to plotting. 

* Boxplot - The Y values must be transformed to the median, first and third quartiles etc. 
* Smoothers - The Y values must be transformed into predicted values. 

## Statistical Transformation

```{r Histogram}
ggplot(midwest, aes(x=area)) + # Only need to set X axis for histogram
  geom_histogram() # Call histogram 
```

## Statistical Transformation
```{r Histogram bin}
ggplot(midwest, aes(x=area)) + # Only need to set X axis for histogram
  geom_histogram(stat = "bin" , binwidth=0.001) # Call histogram 
```

## Changing the Statistical Transformation

```{r bar chart}
pop.sum <- aggregate(midwest["popdensity"], midwest["state"], FUN=mean)
head(pop.sum)

```


## Changing the Statistical Transformation


```{r bar chart 2, eval=FALSE}
ggplot(housing.sum, aes(x=State, y=Home.Value)) + 
  geom_bar()
```

## Changing the Statistical Transformation

```{r fixed bar}
ggplot(pop.sum, aes(x=state, y=popdensity)) + 
  geom_bar(stat="identity")

```


## Exercises 2

1. Create a scatter plot with percent college educated, "percollege", on the X axis and percent below the poverty line, "percbelowpoverty" on Y axis. 
2. Overlay a smoothing line on top of the scatter plot using `geom_smooth` with a glm fit. Hint: see ?geom_smooth
3. Overlay a smoothing line on top of the scatter plot using geom_line. Hint: change the statistical transformation.

## Scales 

* Position
* Color and fill
* Size
* Shape
* Line type


## Common Scale Arguments

* Name - Control the legend or axis title.
* Limits - The minimum and maximum of the scale.
* Breaks - The points along the scale where labels should appear.
* labels - The labels which appear at each break. 


## Scales

```{r scale change example}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() +# Use geom point to add points.
    labs(subtitle="Area Vs Population", 
       title="Scatterplot", 
       caption = "Source: midwest") +
  xlim(0,0.075) +
  ylim(0, 30)
```


## Themes 

The themes in `ggplot2` handles the non-data plot elements, These include,

* Axis Labels
* Plot background
* Facet label background
* Legend appearance

## Black and White Theme

```{r BW Theme}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() +# Use geom point to add points.
  theme_bw()
```


## Classic Theme

```{r Classic Theme}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() +# Use geom point to add points.
  theme_classic()
```

## Nice Plots

```{r Nice Plots}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() +# Use geom point to add points.
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
```

## Facets

Facets are a useful aspect in ggplot which allow us to use a paneled display with the key idea being to display different subsets of the data. These are often referred to as small multiples. There are two options in ggplot for faceting your data:

1. `facet_wrap` - Subsets are defined on a single variable.
2. `facet_grid` - Subsets are defined using two grouping variables. 


## Facets

```{r facet pre}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() # Use geom point to add points.

```


## `facet_wrap`

```{r facet_wrap}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() + # Use geom point to add points. 
  facet_wrap(~state) # Use state for faceting
```

## `facet_grid`

```{r facet_grid}
ggplot(midwest, aes(x=area, y=percbelowpoverty, size=popdensity, color=state)) + # Create ggplot2, x-axis area, y-axis poptotal.
  geom_point() + # Use geom point to add points. 
  facet_wrap(inmetro~state) # Use state for faceting
```


## Putting it all together

Re-create the plot from start of tutorial

Some hints:

1. To remove the confidence bands, you can set `se=F`
2. Set X and Y limits.
3. Check `?labs`

## Plot Code

```{r final plot, warning=FALSE}
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + 
  geom_smooth(method="loess", se=F) + 
  xlim(c(0, 0.1)) + 
  ylim(c(0, 500000)) + 
  labs(subtitle="Area Vs Population", 
       y="Population", 
       x="Area", 
       title="Scatterplot", 
       caption = "Source: midwest") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
plot(gg)
```


## Summary

`ggplot2` is an easy to use but powerful data visualization tool. It allows us to break all our plots down into a set of components and think about plotting in many dimensions for a single graph. 































































































